SFJ: An implementation of Semantic Featherweight Java

### 2020-04-24 conference/camera ready/background.tex
no need for typing
on polygons

\subsection{Types}
\label{sec:types}
The syntax of types $\tau$ is defined by the following grammar:
$$
\begin{array}{llll}
\tau & ::= & \alpha \ |\ \mu
& \mbox{{\em Types}}
\\
\alpha & ::=  & \zero \ |\ \btypes \ |\ [\wt{l:\tau}] \ |\ \alpha\ \synwedge\ \alpha \ |\ \synneg\ \alpha
\qquad
& \mbox{{\em Field types} ($\alpha $-types)}
\\
\mu & ::=  & \alpha \to \alpha \ |\ \mu\ \synwedge\ \mu \ |\ \synneg\ \mu
& \mbox{{\em Method types} ($\mu $-types)}
\end{array}
$$

The $\alpha$-types are used to type fields and the $\mu$-types are used to type methods.
Type $\zero$ is the empty type.
Type $\btypes$ denotes the \emph{basic} types, such as integers, booleans, etc.
{Record} types $[\wt{l:\tau}]$, where $\wt l$ is a sequence of disjoint labels, are used to type objects.
Arrow types $\alpha \to \alpha$ are used to type methods.

@@

\subsection{Terms}
\label{sec:terms}
The syntax of terms is defined by the following grammar and is based on the standard syntax of terms in FJ \cite{featherweight}:
\begin{align*}
     & \mbox{\textit{Class declaration}}  & L \; ::= \; & \syndecl{C}{C}{\wt{\alpha \ a};\ K; \ \wt{M}\ }                         \\
     & \mbox{\textit{Constructor}}        & K \;::=\;   & C\ (\wt{\alpha\ x})\ \{\ \synsuper(\wt{x});\ \wt{\synthis.a}=\wt{x}; \} \\
     & \mbox{\textit{Method declaration}} & M \; ::= \; & \alpha \ m\ (\alpha \ x)\ \{\ \synret e; \}                             \\
     & \mbox{\textit{Expressions}}        & e \; ::=\;  & x\ |\  c\ |\ e.a\ |\ e.m(e) \ | \ \synnew C(\wt{e})
\end{align*}

We assume an infinite set of names, with some special names:
$\mathit{Object}$ denotes the root class, $\synthis$ denotes the current object and $\synsuper$ denotes the parent object.
We let  $A, B,\ldots$ range over classes; $a, b,\ldots$ over fields; $m, n,\ldots$ over methods and $x, y, z, \ldots$ over variables.

A {program} $(\wt{L}, e)$ is a pair of a sequence of class declarations $\wt L$, giving rise to a class hierarchy as specified by the inheritance relation, and an expression $e$ to be evaluated.
%
A class declaration $L$ specifies the name of the class, the name of the parent class it extends, its typed fields, the constructor $K$ and its method declarations $M$.
The constructor $K$ initialises the fields of the object by assigning values to the fields inherited by the super-class and to the fields declared in the current \textbf{this} class.
%
A method declaration $M$ specifies the signature of the method, namely the return type, the method name and the formal parameter as well as the body of the method.
Notice that in our theoretical development we use unary methods, without loss of generality: tuples of arguments can be modelled by an object that instantiates a ``special'' class containing as fields all the needed arguments.
Expressions $e$ include variables, constants, field accesses, method invocations and object creations.

>>>>

\subsection{Types}
\label{sec:types}
The syntax of types $\tau$ is defined by the following grammar:
$$
\begin{array}{llll}
\tau & ::= & \alpha \ |\ \mu
& \mbox{{\em Types}}
\\
\alpha & ::=  & \zero \ |\ \btypes \ |\ [\wt{l:\tau}] \ |\ \alpha\ \synwedge\ \alpha \ |\ \synneg\ \alpha
\qquad
& \mbox{{\em Field types} ($\alpha $-types)}
\\
\mu & ::=  & \alpha \to \alpha \ |\ \mu\ \synwedge\ \mu \ |\ \synneg\ \mu
& \mbox{{\em Method types} ($\mu $-types)}
\end{array}
$$
The $\alpha$-types are used to type fields and the $\mu$-types are used to type methods.
Type $\zero$ is the empty type.
Type $\btypes$ denotes the \emph{basic} types, such as integers, booleans, etc.
{Record} types $[\wt{l:\tau}]$, where $\wt l$ is a sequence of disjoint labels, are used to type objects.
Arrow types $\alpha \to \alpha$ are used to type methods.

@@

\subsection{Terms}
\label{sec:terms}
The syntax of terms is defined by the following grammar and is based on the standard syntax of terms in FJ \cite{featherweight}:
\begin{align*}
     & \mbox{\textit{Class declaration}}  & L \; ::= \; & \syndecl{C}{C}{\wt{\alpha \ a};\ K; \ \wt{M}\ }                         \\
     & \mbox{\textit{Constructor}}        & K \;::=\;   & C\ (\wt{\alpha\ x})\ \{\ \synsuper(\wt{x});\ \wt{\synthis.a}=\wt{x}; \} \\
     & \mbox{\textit{Method declaration}} & M \; ::= \; & \alpha \ m\ (\alpha \ x)\ \{\ \synret e; \}                             \\
     & \mbox{\textit{Expressions}}        & e \; ::=\;  & x\ |\  c\ |\ e.a\ |\ e.m(e) \ | \ \synnew C(\wt{e})
\end{align*}
We assume an infinite set of names, with some special names:
$\mathit{Object}$ denotes the root class, $\synthis$ denotes the current object and $\synsuper$ denotes the parent object.
We let  $A, B,\ldots$ range over classes; $a, b,\ldots$ over fields; $m, n,\ldots$ over methods and $x, y, z, \ldots$ over variables.

A {program} $(\wt{L}, e)$ is a pair of a sequence of class declarations $\wt L$, giving rise to a class hierarchy as specified by the inheritance relation, and an expression $e$ to be evaluated.
%
A class declaration $L$ specifies the name of the class, the name of the parent class it extends, its typed fields, the constructor $K$ and its method declarations $M$.
The constructor $K$ initialises the fields of the object by assigning values to the fields inherited by the super-class and to the fields declared in the current \textbf{this} class.
%
A method declaration $M$ specifies the signature of the method, namely the return type, the method name and the formal parameter as well as the body of the method.
Notice that in our theoretical development we use unary methods, without loss of generality: tuples of arguments can be modelled by an object that instantiates a ``special'' class containing as fields all the needed arguments.
Expressions $e$ include variables, constants, field access, method call and object creation.

### 2020-04-24 conference/camera ready/paper.tex
on polygons

    Consider the set of polygons, such as triangles, squares and rhombuses given by a class hierarchy.
    We want to define a method \emph{diagonal} that takes a polygon and returns the length of its longest diagonal.
    This method makes sense only if the polygon passed to it has at least four sides, hence triangles are excluded.
    In Java this could be implemented in the following ways:

    \begin{equation*}
        \begin{array}{l}
            \synclass Polygon \ \{ \ldots\}
            \\
            \syndecl{Triangle}{Polygon} {\ldots}
            \\
            \syndecl{Other\_Polygons}{Polygon} {
                \\
                \qquad \mathbf{double}\ diagonal(Other\_Polygons \ shape)\ \{\ldots \}
                \\
                \hspace{3.5cm} \ldots
                \\
            }
            \\
            \syndecl{Square}{Other\_Polygons}{\ldots}
            \\
            \syndecl{Rhombus}{Other\_Polygons} {\ldots}
        \end{array}
    \end{equation*}

    Or by means of an interface \emph{Diagonal}:

    \begin{equation*}
        \begin{array}{l}
            public \ \syninterface Diagonal\ \{
            \\
            \qquad \mathbf{double}\ diagonal(Polygon \ shape);
            \\
            \}
            \\
            \synclass Polygon \ \{\ldots\}
            \\
            \syndecl{Triangle}{Polygon} {\ldots}
            \\
            \synidecl{Square}{Polygon} \synimpl Diagonal \ \{\ldots\}
            \\
            \synidecl{Rhombus}{Polygon} \synimpl Diagonal \ \{\ldots\}
            \\
            \hspace{4.5cm} \ldots
        \end{array}
    \end{equation*}

    Now, suppose our class hierarchy is such that \emph{Polygon} is the parent class and all other geometric figures extend \emph{Polygon}, which is how one would naturally define the set of polygons.
    Suppose the class hierarchy is given and is part of legacy code, which cannot be changed.
    Then again, a natural way to implement this in Java is by defining the method \emph{diagonal} in the class \emph{Polygon} and using an \textbf{instanceof}, for example, inside a \textbf{try-catch} construct.
    Then, an exception would be thrown at runtime, if the argument passed to the method is a triangle.

    We propose a more elegant solution, by combining boolean types and semantic subtyping, where only static typechecking is required and we implement this in SFJ \cite{UD20}: it is enough to define a method \emph{diagonal} that has an argument of type \emph{Polygon} \textbf{and not} \emph{Triangle}, thus allowing the typechecker to check at compile time the restrictions on types:

    \begin{equation*}
        \begin{array}{l}
            \synclass Polygon \ \{\ldots\}
            \\
            \syndecl{Triangle}{Polygon} {\ldots}
            \\
            \syndecl{Square}{Polygon} {\ldots}
            \\
            \syndecl{Rhombus}{Polygon} {\ldots}
            \\
            \hspace{2.5cm} \ldots
            \\
            \synclass Diagonal \ \{
            \\
            %\qquad Diagonal ( )\   \{\ super();\ \}
            \hspace{2.5cm} \ldots \\
            \qquad \mathbf{double}\ diagonal((\textit{Polygon} \textbf{ and not }  \textit{Triangle}) \ shape)\{\ldots\}
            \\
            \}
        \end{array}
    \end{equation*}

    We can now call \emph{diagonal} on an argument of type \emph{Polygon}: if the polygon is \textbf{not} a \emph{Triangle}, then the method computes and returns the length of its longest diagonal; otherwise, there will be a type error at compile time.
\end{example}

@@

\section{SFJ: Design and Implementation}
\label{sec:design}
\subsection{Valid type definitions}
Since we want to use types $\tau$ in practice in SFJ, we restrict them to \emph{finite trees} whose leaves are basic types $\btypes$ \autoref{sec:terms} with no cycles.
For example, a recursive type $\alpha = [a : \alpha]$ denotes an infinite program tree \textbf{new} $C$(\textbf{new} $C$($\cdots{}$)), hence we rule it out as it is not inhabitable.
Similarly, the types $\alpha = [b: \beta]$, $\beta = [a = \alpha]$ create a cycle and thus would not be inhabitable.
Notice that these types can be defined and inhabited in Java by assigning \emph{null} to all fields in a class, however they are not useful in practice.

@@

\subsection{Building semantic subtyping for SFJ}

If \autoref{alg:types} returns $True$, meaning all type definitions in a program are valid, we can then build the semantic subtyping.
Leveraging the interpretation of types as sets of values to define semantic subtyping for FJ \cite{Dardha2013,Dardha2017}, in SFJ we keep track of the semantic subtyping relation by defining a map from a type to the set of its subtypes, satisfying the property that the set of values of a subtype is included in the set of values of the type.
We start with basic types and map $Universe$ to the set of all types.
The full mapping for basic types is defined as follows:

\begin{equation}
    \label{eq:initialSubtype}
    \begin{array}{llllll}
        & Double   & = \left\{Double, Float, Int, Short, Byte\right\} &  & Float & = \left\{Float, Short, Byte\right\} \\
        & Long     & =    \left\{Long, Int, Short, Byte\right\}       &  & Int   & = \left\{Int, Short, Byte\right\}   \\
        & Short    & =    \left\{Short, Byte\right\}                  &  & Byte  & = \left\{Byte\right\}               \\
        & Boolean  & =    \left\{Boolean\right\}                      &  & Void  & = \left\{Void\right\}               \\
        & Universe & = \{Double, Float, Long, Int,                                                                     \\
        &          & \quad Short, Byte, Boolean, Void\}
   \end{array}
\end{equation}

Note that \emph{Int} is not a subtype of \emph{Float} as a 32-bit \emph{float} cannot represent the whole set of 32-bit \emph{integer} values accurately and therefore \emph{Int} is not fully set-contained in \emph{Float}, however this is not the case for \emph{Int} and \emph{Double}.
Similarly, \emph{Long} is not a subtype of \emph{Double}.

@@

\subsection{Type system for SFJ}

@@

\subsection{Polygons: continued}

Let us illustrate the subtyping algorithm with our \emph{Polygons} \autoref{polygons}.
The algorithm generates the following subtyping relation, which also includes the subtyping relation for basic types defined in the first step.

\begin{flalign*}
    &Polygon = \left\{Polygon, Triangle, Square, Rhombus \right\} \qquad Triangle = \left\{Triangle\right\}\\
    &Square = \left\{Square\right\} \qquad  Rhombus  = \left\{Rhombus\right\} \qquad
    Diagonal = \left\{Diagonal\right\}
\end{flalign*}

Recall method \emph{diagonal} in class \emph{Diagonal}, we can see that the result of the set operation on its parameter type gives the following set of polygons:

\begin{flalign*}
    Polygon \textbf{ and not } Triangle = \left\{Polygon, Square, Rhombus \right\}
\end{flalign*}

If we pass to \emph{diagonal} an argument of type \emph{Triangle}, as it is not contained in the computed set of subtypes, we get a typechecking error at compile time \cite{UD20}.

%It finds all nominal subtypes as they inherit all fields and methods of their super class, so are always guaranteed to be a superset.
%Similarly, given all pairs of types, the algorithm checks if one is a superset of the other, i.e.,\ they share common fields and methods, and if so, they are related by structural subtyping.
%For example, type $empty = []$, will have all classes as structural subtypes.
%The above algorithm will without differentiating, find all nominal and structural subtypes as we always consider all other classes in the program to be potential subtype candidates.

\subsection{Nominal vs Structural}

>>>>

    Consider the set of polygons, such as triangles, squares and rhombuses given by a class hierarchy.
    We want to define a method \emph{diagonal} that takes a polygon and returns the length of its longest diagonal.
    This method makes sense only if the polygon passed to it has at least four sides, hence triangles are excluded.
    In Java this could be implemented in the following ways:
    \begin{equation*}
        \begin{array}{l}
            \synclass Polygon \ \{ \ldots\}
            \\
            \syndecl{Triangle}{Polygon} {\ldots}
            \\
            \syndecl{Other\_Polygons}{Polygon} {
                \\
                \qquad \mathbf{double}\ diagonal(Other\_Polygons \ shape)\ \{\ldots \}
                \\
                \hspace{3.5cm} \ldots
                \\
            }
            \\
            \syndecl{Square}{Other\_Polygons}{\ldots}
            \\
            \syndecl{Rhombus}{Other\_Polygons} {\ldots}
        \end{array}
    \end{equation*}
    Or by means of an interface \emph{Diagonal}:
    \begin{equation*}
        \begin{array}{l}
            public \ \syninterface Diagonal\ \{
            \\
            \qquad \mathbf{double}\ diagonal(Polygon \ shape);
            \\
            \}
            \\
            \synclass Polygon \ \{\ldots\}
            \\
            \syndecl{Triangle}{Polygon} {\ldots}
            \\
            \synidecl{Square}{Polygon} \synimpl Diagonal \ \{\ldots\}
            \\
            \synidecl{Rhombus}{Polygon} \synimpl Diagonal \ \{\ldots\}
            \\
            \hspace{2.5cm} // \text{ other geometric figures} \ldots
        \end{array}
    \end{equation*}
    Now, suppose our class hierarchy is such that \emph{Polygon} is the parent class and all other geometric figures extend \emph{Polygon}, which is how one would naturally define the set of polygons.
    Suppose the class hierarchy is given and is part of legacy code, which cannot be changed.
    Then again, a natural way to implement this in Java is by defining the method \emph{diagonal} in the class \emph{Polygon} and using an \textbf{instanceof}, for example, inside a \textbf{try-catch} construct.
    Then, an exception would be thrown at runtime, if the argument passed to the method is a triangle.

    We propose a more elegant solution, by combining boolean types and semantic subtyping, where only static typechecking is required and we implement this in SFJ \cite{UD20}: it is enough to define a method \emph{diagonal} that has an argument of type \emph{Polygon} \textbf{and not} \emph{Triangle}, thus allowing the typechecker to check at compile time the restrictions on types:
    \begin{equation*}
        \begin{array}{l}
            \synclass Polygon \ \{\ldots\}
            \\
            \syndecl{Triangle}{Polygon} {\ldots}
            \\
            \syndecl{Square}{Polygon} {\ldots}
            \\
            \syndecl{Rhombus}{Polygon} {\ldots}
            \\
            \hspace{2.5cm} \ldots
            \\
            \synclass Diagonal \ \{
            \\
            %\qquad Diagonal ( )\   \{\ super();\ \}
            \hspace{2.5cm} \ldots \\
            \qquad \mathbf{double}\ diagonal((\textit{Polygon} \textbf{ and not }  \textit{Triangle}) \ shape)\{\ldots\}
            \\
            \}
        \end{array}
    \end{equation*}
    We can now call \emph{diagonal} on an argument of type \emph{Polygon}: if the polygon is \textbf{not} a \emph{Triangle}, then the method computes and returns the length of its longest diagonal; otherwise, there will be a type error at compile time.
\end{example}

@@

\section{SFJ: Design and Implementation}
\label{sec:design}
\subsection{On Valid Type Definitions}
Since we want to use types $\tau$ in practice in SFJ, we restrict them to \emph{finite trees} whose leaves are basic types $\btypes$ \autoref{sec:terms} with no cycles.
For example, a recursive type $\alpha = [a : \alpha]$ denotes an infinite program tree \textbf{new} $C$(\textbf{new} $C$($\cdots{}$)), hence we rule it out as it is not inhabitable.
Similarly, the types $\alpha = [b: \beta]$, $\beta = [a = \alpha]$ create a cycle and thus would not be inhabitable.
Notice that these types can be defined and inhabited in Java by assigning \emph{null} to all fields in a class, however they are not useful in practice.

@@

\subsection{Building Semantic Subtyping for SFJ}

If \autoref{alg:types} returns $True$, meaning all type definitions in a program are valid, we can then build the semantic subtyping.
Leveraging the interpretation of types as sets of values to define semantic subtyping for FJ \cite{Dardha2013,Dardha2017}, in SFJ we keep track of the semantic subtyping relation by defining a map from a type to the set of its subtypes, satisfying the property that the set of values of a subtype is included in the set of values of the type.
We start with basic types and let $Universe$ be a supertype of all types.
The full mapping for basic types is defined in \autoref{eq:initialSubtype}.
\begin{equation}
    \label{eq:initialSubtype}
    \begin{array}{llllll}
        & Double   & = \left\{Double, Float, Int, Short, Byte\right\} &  & Float & = \left\{Float, Short, Byte\right\} \\
        & Long     & =    \left\{Long, Int, Short, Byte\right\}       &  & Int   & = \left\{Int, Short, Byte\right\}   \\
        & Short    & =    \left\{Short, Byte\right\}                  &  & Byte  & = \left\{Byte\right\}               \\
        & Boolean  & =    \left\{Boolean\right\}                      &  & Void  & = \left\{Void\right\}               \\
        & Universe & = \{Double, Float, Long, Int,                                                                     \\
        &          & \quad Short, Byte, Boolean, Void\}
   \end{array}
\end{equation}
Note that \emph{Int} is not a subtype of \emph{Float} as a 32-bit \emph{float} cannot represent the whole set of 32-bit \emph{integer} values accurately and therefore \emph{Int} is not fully set-contained in \emph{Float}, however this is not the case for \emph{Int} and \emph{Double}.
Similarly, \emph{Long} is not a subtype of \emph{Double}.

@@

\subsection{Type System for SFJ}

@@

\subsection{Polygons: Continued}

Let us illustrate the semantic subtyping algorithm on our \emph{Polygons} given in \autoref{polygons}.
\autoref{alg:subtyping} generates the subtyping relation given in \autoref{eq:sugtypePolygons}, together with the subtyping relation for basic types, omitted here and defined in \autoref{eq:initialSubtype}. Notice that the mapping for $Universe$ is extended with the new types for polygons.
\begin{equation}
    \label{eq:sugtypePolygons}
    \begin{array}{llllll}
         & Polygon  & = \left\{Polygon, Triangle, Square, Rhombus \right\} &  & Triangle & = \left\{Triangle\right\}           \\
         & Square   & = \left\{Square\right\}                              &  & Rhombus  & = \left\{Rhombus\right\}            \\
         & Universe & = \{Double, Float, Long, Int, Short, Byte                                                              &  & Diagonal & = \left\{Diagonal\right\}          \\
         &          & \quad Boolean, Void, Polygon, Square                                                                     \\
         &          & \quad Square, Rhombus, Diagonal\}
    \end{array}
\end{equation}
Recall the method \emph{diagonal} in class \emph{Diagonal}, with signature
$$
\mathbf{double}\ diagonal((\textit{Polygon} \textbf{ and not }  \textit{Triangle}) \ shape)
$$
The result of the set operation on its parameter type gives the following set of polygons:
$$
Polygon \textbf{ and not } Triangle = \left\{Polygon, Square, Rhombus \right\}
$$
In order to define the \textbf{not} $Triangle$ type we need the $Universe$ type so that we can define it as $Universe\ \mathbf{\setminus}\ Triangle$. Then, the \textbf{and} connective is the intersection of sets of $Polygon$ with \textbf{not} $Triangle$.

Thus, if we pass to \emph{diagonal} an argument of type \emph{Triangle}, we get a typechecking error at compile time, as illustrated in the accompanying video \cite{UD20}.

\subsection{Nominal vs. Structural Subtyping}
